
---
title: "Dynamic regression models"
output: html_document
date: "2025-08-21"
---


```{r setup, include=FALSE}
library(fpp3)
library(readr)

vic_elec_daily <- vic_elec |>
  filter(year(Time) == 2014) |>
  index_by(Date = date(Time)) |>
  summarise(
    Demand = sum(Demand) / 1e3,
    Temperature = max(Temperature),
    Holiday = any(Holiday)
  ) |>
  mutate(Day_Type = case_when(
    Holiday ~ "Holiday",
    wday(Date) %in% 2:6 ~ "Weekday",
    TRUE ~ "Weekend"
  ))
```


## Dynamic harmonic regression
\fontsize{14}{15}\sf

\textbf{Combine Fourier terms with ARIMA errors}

### Advantages
   * it allows any length seasonality;
   * for data with more than one seasonal period, you can include Fourier terms of different frequencies;
   * the seasonal pattern is smooth for small values of $K$ (but more wiggly seasonality can be handled by increasing $K$);
   * the short-term dynamics are easily handled with a simple ARMA error.

### Disadvantages
 * seasonality is assumed to be fixed

## Eating-out expenditure

```{r cafe, echo=TRUE, fig.height=2.6}
aus_cafe <- aus_retail |>
  filter(Industry == "Cafes, restaurants and takeaway food services",
         year(Month) %in% 2004:2018) |>
  summarise(Turnover = sum(Turnover))
aus_cafe |> autoplot(Turnover)
```

Again we see an increase as the series progresses, and we use the log of the series to counter this.

## Eating-out expenditure

\fontsize{10}{10.5}\sf

```{r cafefit, dependson='cafe', echo=TRUE, results='hide'}
fit <- aus_cafe |> model(
  `K = 1` = ARIMA(log(Turnover) ~ fourier(K = 1) + PDQ(0, 0, 0)),
  `K = 2` = ARIMA(log(Turnover) ~ fourier(K = 2) + PDQ(0, 0, 0)),
  `K = 3` = ARIMA(log(Turnover) ~ fourier(K = 3) + PDQ(0, 0, 0)),
  `K = 4` = ARIMA(log(Turnover) ~ fourier(K = 4) + PDQ(0, 0, 0)),
  `K = 5` = ARIMA(log(Turnover) ~ fourier(K = 5) + PDQ(0, 0, 0)),
  `K = 6` = ARIMA(log(Turnover) ~ fourier(K = 6) + PDQ(0, 0, 0)))
glance(fit)
```
We consider all possible fourier specifications, up to $K=6$. Interestingly, the lowest AICc is given by $K=6$.


```{r, echo = FALSE}
glance(fit) |>
  select(.model, sigma2, log_lik, AIC, AICc, BIC) |>
  knitr::kable()
```

## Eating-out expenditure

```{r, echo=FALSE}
cafe_plot <- function(...) {
  fit |>
    select(...) |>
    forecast() |>
    autoplot(aus_cafe) +
    labs(title = sprintf("Log transformed %s, fourier(K = %s)", model_sum(select(fit, ...)[[1]][[1]]), deparse(..1))) +
    geom_label(
      aes(x = yearmonth("2007 Jan"), y = 4250, label = paste0("AICc = ", format(AICc))),
      data = glance(select(fit, ...))
    ) +
    geom_line(aes(y = .fitted), colour = "red", augment(select(fit, ...))) +
    ylim(c(1500, 5100))
}
```

```{r cafe1, dependson='cafe', fig.height=4, echo=FALSE}
cafe_plot(K = 1)
```

## Eating-out expenditure

```{r cafe2, dependson='cafe', fig.height=4, echo=FALSE}
cafe_plot(K = 2)
```
Notice that the ARIMA term gets simpler for larger values of $K$!

## Eating-out expenditure

```{r cafe3, dependson='cafe', fig.height=4, echo=FALSE}
cafe_plot(K = 3)
```

## Eating-out expenditure

```{r cafe4, dependson='cafe', fig.height=4, echo=FALSE}
cafe_plot(K = 4)
```

## Eating-out expenditure

```{r cafe5, dependson='cafe', fig.height=4, echo=FALSE}
cafe_plot(K = 5)
```

## Eating-out expenditure

```{r cafe6, dependson='cafe', fig.height=4, echo=FALSE}
cafe_plot(K = 6)
```
Note that here we have montly data, which can in fact easily be handled by a Seasonal ARIMA model. We now look at an example where Fourier terms help a lot: An example with \textbf{weekly data} with non-integer seasonality.


## Example: weekly gasoline products
\fontsize{8}{8}\sf

After some trying around we set $K=6$:
```{r gasmodel, echo=TRUE, cache=FALSE}
fit <- us_gasoline |>
  model(K06 = ARIMA(Barrels ~ fourier(K = 6) + PDQ(0, 0, 0)))
report(fit)
```

## Example: weekly gasoline products

```{r gasf, echo=TRUE, fig.height=3.2}
forecast(fit, h = "3 years") |>
  autoplot(us_gasoline)
```

We have quite large prediction intervals, as there is a lot of noise in this dataset, but overall the prediction appears to make sense, which is quite impressive for a 3 year forecast of weekly data.
